Maximum Entropy Spectral Analysis: a case study

### 2021-02-22 paper/mesa.tex
Added some text

\subsection{Choosing an optimizer - A gaussian PSD} \label{sec:arp_validation}
We test the performance of the three optimizers on a random time series with a known power spectral density.
We take the PSD to be a Gaussian distribution with mean $\mu = 2.5 \quad units$ and standard deviation $\sigma = 0.5 \quad units$, see fig.~\ref{fig:autocorr}.
Of course, the time series will be generated by sampling from ${p(\tilde{n}(f_i)) \propto \exp{\left\{-\frac{1}{2}\sum_i\frac{n_i ^2}{S(f_i)}\right\}}}$ and performing the inverse Fast Fourier Transform on the frequency series.

@@

For our studies, we generate a dataset of 1000 time series of 3000 points each: see fig.~\ref{fig:noise} for an example of a time series in frequency and time domain.
For every time series we will compute the spectrum with every optimization method.

\sschmidt{Sono arrivato fino a qui nelle modifiche}

\subsubsection{Final Prediction Error (FPE)}
The estimate of the recursive order obtained with FPE on the 1000 realization of noise, is found to lay between 30 and 70. Qualitatively, there is a good agreement between the reconstructed spectrum and the true spectrum: a logarithmic plot of the mean of all the spectra versus the true spectrum is reported in \textbf{Figure \ref{fig:FPEmean}} with 90\% credible regions. It is evident that, despite not too large error bars, the tail of the normal distribution is not comprised in the interval of the estimation. 
\begin{figure}[H]
    \centering
    \includegraphics[width = \linewidth]{Images/NormalPSD/FPEpsdAndResiduals.pdf}
    \caption{Ensemble Average for FPE reconstruction with 90\% Credibility Regions (above) and percentage error (below)}
    \label{fig:FPEmean}
\end{figure}
To assess the magnitude of the disagreement, we compute the relative error between the mean spectrum $\hat{S}$ and the true spectrum $S$ for every frequency bin
\begin{equation}
    \nonumber
    r(f) = \frac{|\hat{S}(f) - S(f)|}{S(f)}\,.
\end{equation}
We found this to be a better choice than residuals since relative errors are not dependent on the magnitude of the data. Second Fig \ref{fig:FPEmean} (below) shows that the largest error in the reconstruction of the spectrum is in the tails, while in the bulk of the distribution we have a good approximation of the true function. 
For the entire spectrum, the mean relative error $\Bar{r}$ is 
\begin{equation}
    \nonumber
    \bar{r}_{\bar S, FPE} = 0.025.
\end{equation}
FPE performs well in reconstructing the spectrum even with relatively short time series, with a deviation from the true spectrum around 2\%, but it is evident that it is not able to capture all the features of the spectrum, since the reconstruction in the tails is inaccurate. If we compute the average percentage error in three standard deviations from the mean we gain an order of magnitude estimate of the accuracy of the reconstruction, obtaining: 
\begin{equation}
    \nonumber
    \bar{r}_{\bar S, FPE}^{\mu \pm 3\sigma} = 0.005. 
\end{equation}
and we can state the the method is highly performant in this frequency window. 
Since it is not always possible to work with more then one realization for a process, as in observational experiments, it is important to understand how this method behaves when we apply it on a single time series. In that case, the mean error and its standard deviation are: 
\begin{equation}
    \nonumber
    \bar r_{FPE} = 0.16; \quad \sigma_{r} = 0.16\,.
\end{equation}
The average error of a reconstruction based on a single time series is bigger than the one obtained considering the average of all the reconstructions.
Standard deviation of the errors assume small values and 90\% credibility regions (Cr.) in \ref{fig:FPEmean} are tight. This means that if we randomly take two of the reconstructed spectra, we expect their differences to be statistically small. These facts are evidence for FPE to be a relatively stable optimizer. 
We already said that the estimate for filter's length lays in a well defined range of values. The quality of the reconstruction, as noted in Chapter 1, depends on the order of the reconstructed AR process. 
To investigate this idea further, we plot the value of the optimizer for each recursion and the residuals as a function of the estimate for the order of the autoregressive process (\textbf{Figure \ref{fig:FPEErrorOrder}}),

@@

\sschmidt{In questa sezione potremmo mettere i vari confronti con Welch}
\sschmidt{Mettiamo da qualche parte una descrizione dettagliata del codice?}

>>>>

\subsection{Choosing an optimizer - A gaussian PSD} \label{sec:arp_validation}
We test the performance of the three optimizers on a random time series with a known power spectral density.
We take the PSD to be a Gaussian distribution with mean $\mu = 2.5 \quad units$ and standard deviation $\sigma = 0.5 \quad units$, see fig.~\ref{fig:autocorr}.
Of course, the time series will be generated by sampling a frequency vector from ${p(\tilde{n}(f_i)) \propto \exp{\left\{-\frac{1}{2}\sum_i\frac{n_i ^2}{S(f_i)}\right\}}}$ and performing the inverse Fast Fourier Transform on the frequency series.

@@

For our studies, we generate a dataset of 1000 time series of 3000 points each: see fig.~\ref{fig:noise} for an example of a time series in frequency and time domain.
For every time series we will compute the spectrum with every optimization method.
To assess the magnitude of the disagreement, we compute the relative error between the mean spectrum $\hat{S}$ and the true (analytical) spectrum $S$ for every frequency bin:
\begin{equation}
r(f) = \frac{|\hat{S}(f) - S(f)|}{S(f)}\; .
\end{equation}

We report the average values of the relative error for the different optimizers in Table~\ref{tab:relative_errors}.

\sschmidt{Sono arrivato fino a qui nelle modifiche}
\begin{table}
	\caption{Average relative error for different optimizers, computed in different ways. We consider a set of $1000$ realization of a noisy time series of $3000$ points.
	We report the average error  $\bar{r}_{\bar{S}}$ on the whole set of simulations, the error $\bar{r}_{\bar S}^{\mu \pm 3\sigma} $ averaged only on data points between $3\sigma$ and the error $\bar{r}$ on a single, randomly drawn realization.
	\sschmidt{I think we need to discuss on which performance error to include in the estimation}
	}
	\label{tab:relative_errors}
	\def\arraystretch{1.5}
	\begin{ruledtabular}
		\begin{tabular}{ l c c c }
			\multirow{2}{*}{Optimizer}& \multicolumn{3}{c}{Relative Error}\\
			& $\bar{r}_{\bar{S}}$	& $\bar{r}_{\bar S}^{\mu \pm 3\sigma} $ & $\bar{r}$\\
			\hline \hline
			FPE 			& $6.9 \; (46.9\%)$	& $7 \; (1.6\%)$ & \\ 
			\hline
			CAT 			& $4.5 \; (30.6\%)$	& $194 \; (45.3\%)$ & \\
			\hline
			OBD 				& $1.7 \; (11.6\%)$	& $206 \; (48.1\%)$ & \\
		\end{tabular}
	\end{ruledtabular}
\end{table}

\begin{figure}[H]
	\centering
	\includegraphics[width = \linewidth]{Images/NormalPSD/FPEpsdAndResiduals.pdf}
	\caption{Ensemble Average for FPE reconstruction with 90\% Credibility Regions (above) and percentage error (below)}
	\label{fig:FPEmean}
\end{figure}

\subsubsection{Final Prediction Error (FPE)}
The results of our investigation on te performance of FPE are reported in fig.~\ref{fig:FPEmean}: we plot the ``true" spectrum, the reconstructed spectrum as well as the $90\%$ credible region of the predicted spectrum.
Qualitatively, there is a good agreement between the reconstructed spectrum and the true spectrum; however, we note that the reconstruction is not very accurate in the low frequency region.

For the entire spectrum, the mean relative error $\bar{r}$ is 
\begin{equation}  
    \bar{r}_{\bar S, FPE} = 0.025.
\end{equation}
FPE performs well in reconstructing the spectrum even with relatively short time series, with a deviation from the true spectrum around 2\%, but it is evident that it is not able to capture all the features of the spectrum, since the reconstruction in the tails is inaccurate. If we compute the average percentage error in three standard deviations from the mean we gain an order of magnitude estimate of the accuracy of the reconstruction, obtaining: 
\begin{equation}
    \nonumber
    \bar{r}_{\bar S, FPE}^{\mu \pm 3\sigma} = 0.005. 
\end{equation}
and we can state the the method is highly performant in this frequency window. 
Since it is not always possible to work with more then one realization for a process, as in observational experiments, it is important to understand how this method behaves when we apply it on a single time series. In that case, the mean error and its standard deviation are: 
\begin{equation}
    \nonumber
    \bar r_{FPE} = 0.16; \quad \sigma_{r} = 0.16\,.
\end{equation}
The average error of a reconstruction based on a single time series is bigger than the one obtained considering the average of all the reconstructions.
Standard deviation of the errors assume small values and 90\% credibility regions (Cr.) in \ref{fig:FPEmean} are tight. This means that if we randomly take two of the reconstructed spectra, we expect their differences to be statistically small. These facts are evidence for FPE to be a relatively stable optimizer. 
We already said that the estimate for filter's length lays in a well defined range of values. The quality of the reconstruction, as noted in Chapter 1, depends on the order of the reconstructed AR process. 
To investigate this idea further, we plot the value of the optimizer for each recursion and the residuals as a function of the estimate for the order of the autoregressive process (\textbf{Figure \ref{fig:FPEErrorOrder}}),

@@

It is really interesting to have a qualitative comparison between the performance of the MESA and of the standard Welch algorithm.
We perform such comparison on the spectrum of LIGO (even though similar conclusions can be drawn from any other shape of the PSD).

We perform two sets of experiments.
In the first one (see fig.~), we simulate data from an analytical PSD and we make the comparison with them: this is to endure the we have a baseline PSD to compare the data with.
In a second experiment (see fig.~), we employ the data released to the public byt the LIGO/Virgo collaboration.
In each experiment, we vary the length of the data used for the estimation: this is also useful to assess how the computation depends on the data available. We set the total observation time $T = 1, 5, 10, 100, 1000 \SI{}{s}$
For the MESA algorithm, we employ FPE optimizer. For the Welch algorithm, we employ a Tukey window, an overlap fraction of $0.5$ for the segments and a length of segments $L = 512, 1024, 2048, 8192, 32768$ datapoints.
In all cases, the sampling rate is set to $\SI{4096}{Hz}$.

First of all, we note that using a longer time series results in a more stable estimation of the PSD, especially in the low frequency sector. This is somehow obvious: more data provide more information available for spectral estimation. This fact is more important at lower frequencies, where the natural time length is large.
Turning to the comparison between MESA and Welch's method, we find that the PSD estimation provided by the Welch's method is more noisy (i.e. has a large number of spurious peaks) rather than PSD provided by MESA. The different behavior is more manifest in the low frequency sector.
The result does not change much by changing the hyperparameters required for Welch algorithm.
\sschmidt{Other observations??}

\begin{figure}
	\label{fig:mem_welch_realdata}
	\caption{Blablabla}
\end{figure}

\sschmidt{Mettiamo da qualche parte una descrizione dettagliata del codice?}
